Tutorial Proposal: Writing Interpreters for Dynamic Languages Using PyPy (and Getting Them Fast)

### 2009-01-19 talk/ecoop2009/tutorial/proposal.tex
(antocuni, arigo, cfbolz): write an exciting abstract.

\institute{DISI, University of Genova, Italy 
\and 
Softwaretechnik und Programmiersprachen
 Heinrich-Heine-Universit\"at D\"usseldorf}

\maketitle

The easiest way to implement a dynamic language such as JavaScript or Python is
to write an interpreter for it; however, interpreters are slow. An alternative
is to write a compiler; writing a compiler that targets a high level virtual
machine like CLI or JVM is easier than targeting a real CPU, but it still
require a lot of work, as IronPython, Jython, JRuby demonstrate. Moreover,
the various platforms have to be targeted independently.

The PyPy project\footnote{\url{http://codespeak.net/pypy}} \cite{RiBo07_223}
aims to make the implementation of dynamic
languages easier by providing a toolchain that allows to translate an
interpreter to various target platforms, including the JVM, .NET and C/Posix. 
In addition, we are working on being able to automatically transform the
interpreter into a specializing JIT-compiler to vastly increase performance.
The goal is to minimize the effort required to get a fast implementation of a
dynamic language.

This tutorial will give an insight into the usage of the PyPy translation
toolchain by giving a step-by-step introduction. The topics covered are how to
write an interpreter for a dynamic language for PyPy and make it possible to
automatically generate a JIT for it.  We will showcase a typical interpreter for a
small dynamic language as a running example.

>>>>

\institute{DISI, University of Genova, Italy 
\and 
Softwaretechnik und Programmiersprachen
 Heinrich-Heine-Universit\"at D\"usseldorf}
\maketitle

\begin{abstract}
We propose to give a tutorial on how to use the PyPy project to write
interpreters for dynamic languages. PyPy supports compiling such interpreters
to various target environments (like .NET or C/Posix) and automatically
generating a Just-In-Time compiler for them. PyPy is thus an emerging tool to
help implement fast, portable multi-platform interpreters with less effort.
\end{abstract}

The easiest way to implement a dynamic language such as JavaScript or Python is
to write an interpreter for it; however, interpreters are slow. An alternative
is to write a compiler; writing a compiler that targets a high level virtual
machine like CLI or JVM is easier than targeting a real CPU, but it still
require a lot of work, as demonstrated by the IronPython, Jython, JRuby
projects. Moreover, the various platforms (either VM or real hardware) have to
be targeted independently.

The PyPy project\footnote{\url{http://codespeak.net/pypy}} \cite{RigoPedroni06}
\cite{RiBo07_223}
aims to make the implementation of dynamic
languages easier by providing a toolchain that allows to compile an
interpreter to various target platforms, including the JVM, .NET and C/Posix. 
In addition, we are working automatically transforming the
interpreter into a specializing JIT-compiler to vastly increase performance.
The goal is to minimize the effort required to get a fast implementation for a
dynamic language.

This tutorial will provide insights into the usage of the PyPy translation
toolchain by giving a step-by-step introduction. The topics covered are how to
write an interpreter for a dynamic language with PyPy and make it possible to
automatically generate a JIT for it.  We will showcase a typical interpreter for a
small dynamic language as a running example.
